---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First, load prerequisite packages ggplot2 and dplyr. You must have these installed to run the program.
```{r, results="hide"}
library(dplyr)
library(ggplot2)
```

This code loads the dataset and adds performs some operations to make later analyses easier. I reformat the
Date variable, created a weekday/weekend indicator, and drop unneeded variables.
```{r}
setwd("C:/Users/duncans/Dropbox/Programming/R_WDir")
dta <- read.csv("activity.csv")
dta <- dta %>%
        mutate(Date = as.Date(as.character(date)),
                Weekday = weekdays(Date), 
                Weekend = ifelse(Weekday == "Saturday" | Weekday == "Sunday", 1, 0)) %>%
        select(steps, interval, Date, Weekday, Weekend)
dta$interval <- as.factor(dta$interval)
```

## What is mean total number of steps taken per day?

For brevity here, I report mean number of steps per the first five days. Dataset dta_sums has this information
for all days.
```{r}
dta_sums <- dta %>% 
        na.omit() %>%
        group_by(Date) %>%
        summarize(steps = sum(steps))
head(dta_sums)
```

And here is a histogram showing the mean number of steps per day:
```{r}
hist(dta_sums$steps, main = "Daily Steps", xlab = "Steps")
```

These are the mean and median steps per day (across all days):
```{r}
mean(dta_sums$steps)
median(dta_sums$steps)
```

## What is the average daily activity pattern?

A line plot shows this information below. Each point represents the mean number of steps taken during a time interval.
```{r}
dta_patt <- dta %>% 
        na.omit() %>%
        group_by(interval) %>%
        summarize(MeanSteps = mean(steps))
plot(as.numeric(as.character(dta_patt$interval)), dta_patt$MeanSteps, 
        type = 'l', 
        main = "Average Steps by Time of Day",
        xlab = "Millitary Time", ylab = "Average Steps")
```

This is the five minute interval with the most steps:
```{r}
filter(dta_patt, MeanSteps == max(dta_patt$MeanSteps))
```


## Imputing missing values

I'll begin by counting the number of rows with missing values:
```{r}
dta_missing <- mutate(dta, MissingVar = complete.cases(dta))
nrow(filter(dta_missing, MissingVar == FALSE))
```

I choose to attach the mean step count (for a given interval) to instances with missing step information.
The code below accomplishes this task.
```{r}
dta_editmiss <- merge(x = dta_missing, y = dta_patt,
        by.x = "interval", by.y = "interval", all=TRUE)
dta_editmiss2 <- dta_editmiss %>% 
        mutate(StepsEdit = ifelse(MissingVar == FALSE, MeanSteps, steps)) %>%
        arrange(Date, interval) %>%
        select(StepsEdit, interval, Date, Weekday, Weekend)
```


## Are there differences in activity patterns between weekdays and weekends?

I will create a colored line plot to answer this question. Note that I have already created weekend and weekday
indicator variables. The code below performs a few operations that are necessary prior to making the plot.
```{r}
dta_finalplot <- dta_editmiss2 %>%
        group_by(interval, Weekend) %>%
        summarize(StepsEdit = mean(StepsEdit))
dta_finalplotEnd <- filter(dta_finalplot, Weekend == 1)
dta_finalplotDay <- filter(dta_finalplot, Weekend == 0)
dta_finalplot2 <- merge(dta_finalplotEnd, dta_finalplotDay,
        by.x = "interval", by.y = "interval")
dta_finalplot3 <- dta_finalplot2 %>%
        rename(WeekEndSteps = StepsEdit.x, WeekDaySteps = StepsEdit.y) %>%
        mutate(Interval = as.numeric(as.character(interval))) %>%
        select(Interval, WeekEndSteps, WeekDaySteps) %>%
        arrange(Interval)
```

Create a colored line plot using the ggplot2 graphics package.
```{r}
g <- ggplot(data=dta_finalplot3,
        mapping = aes(Interval, WeekDaySteps)) +
        ggtitle("Steps: Comparing Weekdays to Weekends") +
        xlab("Millitary Time") +
        ylab("Steps") +
        geom_line(color="Blue") +
        geom_line(mapping=aes(Interval, WeekEndSteps), color="Orange") +
        annotate("text", x=1500, y=225, color="Blue", label = "Weekday Steps") +
        annotate("text", x=1750, y=210, color="Orange", label = "Weekend Steps")
g
```

